# Exploring R Data Structures

One of R's most powerful features is its ability to deal with tabular data - such as you may already have in a spreadsheet or a CSV file.

Let's start by creating a toy dataset of `feline-data.csv` in `data/` directory (folder)

But if `data/` directory does not exist, create it first.

Start with orienting yourself in the R environment. Where am I?

```{r}
getwd() # prints the current working directory "where am I?"
dir.create("data") # creates a new directory called "data"
setwd("C:/Users/rhpar/Documents/data") # sets the working directory to "data"
```
::: {.callout-important}
Check if all sticky notes are green
:::

Now, we are inside the `data/` directory. We can now create a toy dataset of feline data and save it as `feline-data.csv`.

First, let's create a data frame in R:

```{r}
cats <- data.frame(coat = c("calico", "black", "tabby"),
                    weight = c(2.1, 5.0, 3.2),
                    likes_catnip = c(1, 0, 1))
```
We can now save `cats` as a CSV file using the `write.csv()` function:
```{r}
write.csv(cats, "feline-data.csv", row.names = FALSE) # it is important to call the argument row.names = FALSE to avoid writing row numbers to the CSV file
```
The contents of the `feline-data.csv` file will look like this:
```{r}
coat,weight,likes_catnip
calico,2.1,1
black,5.0,0
tabby,3.2,1
```
```{r}
cats <- read.csv(file = "data/feline-data.csv")
cats
```
`read.csv()` is a function that reads a CSV file or **comma-separated values** file and creates a data frame in R. The `file` argument specifies the path to the CSV file. In this case, we are reading the `feline-data.csv` file from the `data/` directory.

You can check out other function like `read.delim` for **tab-separated values** file but for now, we will stick with `read.csv`.

::: {.callout-important}
Check if all sticky notes are green
:::

::: {.callout-important}
Check your data for factors
:::

```{r}
str(cats)
```
We should only see numeric and character data types. If you see factors, we can prevent R from automatically creating “factor” data by using the `options(stringsAsFactors = FALSE)` function.

Then re-read the `feline-data.csv` file for the changes to take effect.

We can beging exploring our dataset right away, pulling out columns by specifying them using the `$` operator:

```{r}
cats$weight # prints the weight column

cats$coat # prints the coat column

```

We can also do operation on the columns:
```{r}
## Say we discovered that the scale weigh is two Kg lighter then we have to add 2 Kg to the values in weight column.
cats$weight + 2 # does not change the original data frame, but prints the new values
```

::: {.callout-important}

## Not included in the guide
```{r}
cats$weight <- cats$weight + 2 # changes the original data frame by adding 2 Kg to the values in weight column
```
:::

```{r}
paste("My cat is", cats$coat)
```
But what about
```{r}
cats$weight + cats$coat
```
```{r}
Error in `cats$weight + cats$coat`:
! non-numeric argument to binary operator
```
Understanding what happened here is key to successfully analyzing data in R.

::: {.callout-important}
Any guesses why there is an error message? What do you think the error message means?
:::

The error message is telling us that we cannot add a numeric vector (the weight column) to a character vector (the coat column). This is because R does not know how to combine these two different types of data.



